Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2019-03-15 papers/histogram/sad.tex

Wang and Landau introduced an algorithm (WL) that uses an update
factor and a statistical histogram to compute the density of states of
a given system~\cite{wang2001determining, wang2001efficient}.  While
the method is incredibly powerful, it has a few disadvantage.
Firstly, it requires the user to select a number of parameters,
including the range of energies to be studied~\cite{landau2004new}.
This adds an additional hurdle to its application to a variety of
systems.  The simulation violates detailed balance, with the size of
the violation decreasing with time, which complicates convergence
analysis.  In fact, the error in a WL computation the error has been
demonstrated to saturate at a non-zero value, i.e. the method does not
converge to the true density of states~\davidsays{Find as many
  citations as we can! Presumably some from the following paragraph?}.

@@

\section{SAD Algorithm}\label{sec:sad}
The Stochastic Approximation with Dynamic update factor (SAD) method
is a variant of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. Each
Flat-Histogram method has unique advantages and disadvantages.
Wang-Landau requires an energy range for initialization.  SAMC removes
this energy range requirement but requires simulating every possible
energy. Our proposed method SAD requires the user to input
$T_\text{min}$ which is an immediate disadvantage of the method;
however, this allows the user to define an energy simulation
range between $T_\text{min}$ and $\infty$. Setting a physical parameter
such as a minimum temperature $T_\text{min}$ is intuitive vs. a user
needing to know a prior either an energy range or some unphysical
parameter $t_0$.

@@

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
the SAMC parameter
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy (see Fig.~\ref{fig:entropy-cartoon}), this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by
$E_H -E_L$ where $E_H$ and $E_L$ are defined below.
The entropy numerator of the update factor
in general should scale with the total number of interesting
energy states $N_S$, since updates to the weights are distributed between
that many energy states.  The product $N_S\langle S\rangle$ is the
total change of entropy required (starting from constant weights) to find
the true entropy, and puts a lower bound on the convergence time.
After \emph{long} times, when
all the energies have been long ago found,
we wish for a lower $t_0$ in order to more rapidly refine the remaining
error in entropy.  We enable this
by tracking the last time we encountered a new energy $t_L$, and
gradually transition to lower $t_0$ (but still asymptotically scaling
as $\gamma_t \propto 1/t$ to ensure eventual convergence).  Finally,
we wish for an update factor that is \emph{never} greater than 1, because
a very large update factor could introduce errors in entropy that may take
many iterations to remove.  The SAD
expression for $\gamma_t$ which incorporates these ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is found to be important
during the course of
the simulation, the update factor will experience a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly.

@@

\begin{figure*}
  \includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-default.pdf}%
\includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-slow.pdf}
  \caption{(a) The entropy error for each MC method for $N=50$,
               $\delta_0 = 0.05\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run.  The error is
               averaged over 8 independent simulations, and the best
               and worst simulations for each method are shown as a
               semi-transparent shaded area.
           (b) The update factor $\gamma_t$ versus iteration number
               for the same simulations.
           (c) The entropy error for each MC method for $N=50$,
               $\delta_0 = 0.005\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run. The error is
               determined exactly as in (a).  
           (d) The update factor $\gamma_t$ versus iteration number
               for the same simulations.
  }
\end{figure*}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_50-ff0_17-N256-entropy-error-default.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.17$ and $T_{\min} = 1$ with
  SAD serving as the reference.}\label{fig:n256}
\end{figure}
%\begin{figure}
%  \includegraphics[width=\columnwidth]{figs/N256-lndos-comparison}
%  \caption{\davidsays{This could be prettier, and do we need it? We
%      definitely need a caption if we keep it.} Use this to create
%    description of the system.}\label{fig:n256-lndos-comparison}
%\end{figure}

>>>>

Wang and Landau introduced an algorithm (WL) that uses an update
factor and a statistical histogram to compute the density of states of
a given system~\cite{wang2001determining, wang2001efficient}.  While
the method is incredibly powerful, it has a few disadvantage.
Firstly, it requires the user to select a number of parameters,
including the range of energies to be studied~\cite{landau2004new}.
This adds an additional hurdle to its application to a variety of
systems.  The simulation violates detailed balance, with the size of
the violation decreasing with time, which complicates convergence
analysis.  In fact, the error in a WL computation the error has been
demonstrated to saturate at a non-zero value~\cite{yan2003fast}, i.e. the method does not
converge to the true density of states~\cite{belardinelli2007wang,
belardinelli2008analysis, belardinelli2014intrinsic,
singh2012density}.

@@

\section{SAD Algorithm}\label{sec:sad}
The Stochastic Approximation with Dynamic update factor (SAD) method
is a variant of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. Each
Flat-Histogram method has unique advantages and disadvantages.
Wang-Landau requires an energy range for initialization.  SAMC removes
this energy range requirement but requires simulating every possible
energy. Our proposed method SAD requires the user to input
$T_\text{min}$, the lowest temperature of interest,
which is an immediate disadvantage of the method. However,
setting a physical parameter such as a minimum temperature
$T_\text{min}$ is possibly easier for a user
than determining in advance either an energy range or some unphysical
parameter $t_0$.

@@

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
the SAMC parameter
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy (see Fig.~\ref{fig:entropy-cartoon}), this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by
$E_H -E_L$ where $E_H$ and $E_L$ are defined below.
The entropy numerator of the update factor
in general should scale with the total number of interesting
energy states $N_S$, since updates to the weights are distributed between
that many energy states.  The product $N_S\langle S\rangle$ is the
total change of entropy required (starting from constant weights) to find
the true entropy, and puts a lower bound on the convergence time.
After \emph{long} times, when
all the energies have been long ago found,
we wish for a lower update factor in order to more rapidly refine the remaining
error in entropy.  We track the time at which we first visted each possible
energy.  We define $t_L$ to be the last time that we encountered an energy
that we currently believe is in the energy range of interest, so a $t\gg t_L$
we feel confident that we have established the true energy range of interest. We
gradually transition to a lower update factor (but still asymptotically scaling
as $\gamma_t \propto 1/t$ to ensure eventual convergence).  Finally,
we wish for an update factor that is \emph{never} greater than 1, because
a very large update factor could introduce errors in entropy that may take
many iterations to remove.  The SAD
expression for $\gamma_t$ which incorporates these ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is found to be important
during the course of
the simulation, the update factor will experience a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly.

@@

\begin{figure*}
  \includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-default.pdf}%
\includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-slow.pdf}
  \caption{(a) The average entropy error for each MC method for $N=50$,
               $\delta_0 = 0.05\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run.  The error is
               averaged over 8 independent simulations, and the best
               and worst simulations for each method are shown as a
               semi-transparent shaded area, and
           (b) the update factor $\gamma_t$ versus iteration number
               for the same simulations.
           (c) The average entropy error for each MC method for $N=50$,
               $\delta_0 = 0.005\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run, and
           (d) the update factor $\gamma_t$ versus iteration number
               for the same simulations.
  }\label{fig:fast-slow-gamma}
\end{figure*}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_50-ff0_17-N256-entropy-error-default.pdf}
  \caption{
  The average entropy error for each MC method for $N=256$,
               $\delta_0 = 0.05\sigma$, $\eta = 0.17$, and $T_{\min} = 1$
               as a function of number of iterations run.  The error is
               averaged over 8 independent simulations, and the best
               and worst simulations for each method are shown as a
               semi-transparent shaded area.  The update factor for this
               system is in Fig.~\ref{fig:gamma-vs-t} above.}\label{fig:n256}
\end{figure}
%\begin{figure}
%  \includegraphics[width=\columnwidth]{figs/N256-lndos-comparison}
%  \caption{\davidsays{This could be prettier, and do we need it? We
%      definitely need a caption if we keep it.} Use this to create
%    description of the system.}\label{fig:n256-lndos-comparison}
%\end{figure}
